---
title: "Capstone Project Modeling"
subtitle: "Introduction to Data Science"
author: "Kevin Tajkowski"
output: word_document
---

### Library Calls
Load the necessary packages.

```{r Library, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```


### Data Preparation
Load the data into a data frame.  Remove all non-fish species by Family_Name.  Non-fish species observed in the data set include sharks, rays, eels, turtles and a dolphin.

```{r Preparation, message = FALSE}
Little_Cayman <- read_csv("Little_Cayman.csv")
non_fish_family <- c("Requiem Shark", "Conger", "Dolphin", "Stingray", "Manta", "Moray", "Eagle Ray", "Snake Eel", "Carpet Shark", "Hammerhead Shark", "Electric Ray", "Round Stingrays", "Sea Turtles", "#N/A")
LC <- Little_Cayman %>% filter(!(Family_Name %in% non_fish_family))
```


### Data Frame Manipulation
Red lionfish are removed from the data set.  Fish families with less than a prescribed number of surveys for all species in the family are identified and removed from the data set.

```{r Survey Count, message=FALSE, echo=FALSE}
LC <- LC %>% filter(!(English_Name == "Red Lionfish (aka Common Lionfish)")) # Remove all lionfish surveys

count_list <- LC %>% group_by(Family_Name) %>% summarize(count = n()) # Survey count for each family
count_list <- count_list %>% filter(count_list$count > 210) # Remove families with too few surveys
family_list <- as.vector(count_list[['Family_Name']]) # Create a vector list of family names
LC_Mod <- LC %>% filter(Family_Name %in% family_list) # Create new data frame with list
```


### Density Comparison
'LC_Mod' excludes non-fish species, lionfish and fish families with < a certain number of surveys
'family_list' is a vector of family names
This section of code generates two vectors of numerical values.  The values are the slopes of trendlines that fit the abundance versus date plots.  Two plots were generated for each family: before the first known lionfish sighting and after the first sighting.  The vector 'pre_slope' corresponds to the plots of abundance versus date before the lionfish and the vector 'post_slope' corresponds to the plots after the lionfish sighting.

```{r Density Determination, echo=FALSE, message=FALSE}
pre_slope <- NULL
post_slope <- NULL

for(i in seq_along(family_list)){
  pre_LF <- LC_Mod %>% filter(Family_Name == family_list[i]) %>% filter(Date < "2009-02-17")
  pre_m <- lm(pre_LF$Abundance~pre_LF$Date)
  pre_slope[i] <- pre_m$coefficients["pre_LF$Date"]
  
  post_LF <- LC_Mod %>% filter(Family_Name == family_list[i]) %>% filter(Date >= "2009-02-17")
  post_m <- lm(post_LF$Abundance~post_LF$Date)
  post_slope[i] <- post_m$coefficients["post_LF$Date"]
}
```

This section of code is used to generate a density plot of the pre_slope and post_slope vectors created above.
```{r, Density Plotting, message=FALSE, echo=FALSE}
d1 <- density(pre_slope)
d2 <- density(post_slope)
xlim <- range(d1$x, d2$x)
ylim <- range(d1$y, d2$y)
cols <- c("red", "green")
plot(1, xlim=xlim, ylim=ylim, type='n', main='Pre and Post-Lionfish Density Comparison', xlab='Slope of Trendline fit to Family Abundance Data', ylab='Density')
lines(d1$x, d1$y, col=cols[1])
lines(d2$x, d2$y, col=cols[2])
lg <- NULL
lg <- c(lg, sprintf('Pre Lionfish, n=%s', length(pre_slope)))
lg <- c(lg, sprintf('Post Lionfish, n=%s', length(post_slope)))
legend('topleft', lg, lwd=2, col=cols)
```


### Moving Window Average Slope
Use a moving window to analyze the data from specific time frames progressing from the first survey date (March 27, 1994) to the last survey date available (April 17, 2018).  The size of the window (number of years) and the time to shift the window (number of years) are variables.

The following code is used to identify important dates and initialize several variables.
```{r, Initialization, message=FALSE}
initial_date <- as.Date("1994-03-27") # Date that the first survey was taken
final_date <- as.Date("2018-04-17") # Date of the last survey taken
shift <- 1 # Duration from beginning of window to the beginning of next window (in years)
window <- 3 # Size (length of time) of window (in years) - Should be equal to or larger than 'shift'

begin_date <- initial_date # initialize beginning of window to date of first survey
end_date <- begin_date %m+% years(window) # initialize end of window

slope_DF <- data.frame(date = as.Date(NA), mean_slope = NA)

i <- 1
j <- 1
```

The following code is used to determine the average slope of all of the slopes of the trendlines for every family during a specific time period (the width of the window).
```{r, Density Series, message=FALSE, fig.height=3.5}
while(end_date <= final_date){
  current_window <- LC_Mod %>% filter(Date >= begin_date & Date < end_date)
  slopes <- NA
  
  for(i in seq_along(family_list)){
    c_wind <- current_window %>% filter(Family_Name == family_list[i])
    TL_slope <- lm(c_wind$Abundance ~ c_wind$Date)
    slopes[i] <- TL_slope$coefficients["c_wind$Date"]
  }

  slope_DF[[j,1]] <- begin_date
  slope_DF[[j,2]] <- mean(slopes)
  j <- j+1
  
  d1 <- density(slopes)
  xlim <- range(d1$x)
  ylim <- range(d1$y)
  cols <- "red"
  
  plot(1, xlim=xlim, ylim=ylim, type='n', main=paste(begin_date, "to", end_date, "\nWindow length:", window, "years"), xlab='Slope of Trendline fit to Family Abundance Data', ylab='Density')
  rug(slopes)
  density(slopes, adjust = 0.5)
  lines(d1$x, d1$y, col=cols)
  
  begin_date <- begin_date %m+% years(shift)
  end_date <- begin_date %m+% years(window)
}
```

Generate a plot of the average of the slopes for the moving window.
```{r, Mean Values, message=FALSE, fig.height=3.5}

first_LF <- as.Date("2009-02-17")

ggplot(slope_DF, aes(x = date, y = mean_slope)) +
  geom_point(size = 0.75) +
  geom_line() +
  ggtitle("Mean Slope of Trendlines for all Families") +
  theme(plot.title = element_text(hjust = 0.5, size = 12.0)) +
  geom_line(aes(slope_DF$date, 0), col = "green") +
  geom_vline(xintercept=first_LF, col = "red") +
  xlab(paste("Start Date of", window, "- year Window")) +
  ylab("Mean Slope") +
  theme_bw() +
  annotate("text", x = as.Date("2012-01-01"), y = 1.25e-04, label = "First Lionfish", col = "red") +
  annotate("text", x = as.Date("2012-01-01"), y = 0.9e-04, label = "Sighting", col = "red")

```

