---
title: "Capstone Project Site Assessment"
subtitle: "Introduction to Data Science"
author: "Kevin Tajkowski"
output: word_document
---


### Reading the dive site survey file
First I load the necessary package.  I then read in the file with the number and type of surveys per dive site in Little Cayman.  Unnecessary rows at the top are stripped and the columns are properly labeled.  SA is Species and Abundance, SO is Species Only and BT is Bottom Time.  The two levels of surveyors are Expert and Novice.
```{r, message=FALSE}
library(readr)

headings <- c("Site_ID", "Site_Name", "Expert_SA", "Expert_SO", "Novice_SA", "Novice_SO", "BT")
mydata <- read_csv(file="/Users/Kevin/Documents/Dive Site Data.csv", skip = 7, col_names = headings)
```

### Modifiying the data frame
Here I delete the last row and choose the desired columns.  I then create a new column to sum the number of SA (Species and Abundance) dives on each dive site.
```{r}
dive_sites <- mydata[-85,c("Site_ID", "Site_Name", "Expert_SA", "Novice_SA")]
dive_sites$Total_SA <- as.numeric(dive_sites$Expert_SA) + as.numeric(dive_sites$Novice_SA)
```

***

### Plotting
Here I determine the dive site with the greatest number of surveys.  I can then choose the best rnage for the x-axis and plot a histogram.
```{r}
max(dive_sites$Total_SA)

mean(dive_sites$Total_SA)

median(dive_sites$Total_SA)

hist(dive_sites$Total_SA, breaks = seq(0, 300, 25), col="blue", main = "Little Cayman\n Dive Site Survey Count\n (1993 to present)", xlab="Number of Surveys", ylab = "Number of Dive Sites")
abline(v=c(mean(dive_sites$Total_SA), median(dive_sites$Total_SA)), lty=c(2,3), lwd=2)
legend("topright", legend=c("mean Total_SA", "median Total_SA"), lty = c(2,3), lwd=2)

```

### Analysis
Here, I find all of the dive sites with 100 or more surveys, sort the data frame and pick the top five highest surveyed sites.
```{r}
final_sites <- dive_sites[dive_sites$Total_SA>=100,]
final_sites <- final_sites[order(final_sites$Total_SA, decreasing=TRUE), ]
study_sites <- final_sites[c(1:5),]
study_sites[ , c(1,2,5)]
```

***



